---
title: "Reproducible Research Assignment 2"
author: "M. Bruzza"
date: "April 16, 2018"
output: 
  html_document: 
    keep_md: yes
---

## Introduction/Synopsis

## Data Processing

Load all the R packages we plan on using into the current R session

```{r, message = FALSE, warning = FALSE}
# Load required packages into current R session
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(R.utils)
```

### Download and unzip the raw data file
First we'll check to see if we already have the unzipped data file in our
working directory. If not, we will check to see if we have the zipped file, and 
download it if necessary, and then unzip the file.

```{r}
# Check if file exists in working directory
if(!file.exists("storm_data.csv")){
        
        # Download zipped file if needed
        if(!file.exists("zipped_storm_data.csv.bz2")){
                URL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
                download.file(URL,
                              destfile = "zipped_storm_data.csv.bz2")
                rm(URL)
        }
        
        # Unzip data file
        bunzip2(filename = "zipped_storm_data.csv.bz2", 
                destname = "storm_data.csv", 
                remove = F, 
                skip = T)
        }
```

### Read and transform Data
Next we read the data into R. We will group the data based on the type of 
weather event.

```{r, message = FALSE, cache = TRUE}
# Read unzipped data file into current R session
storm_data <- read_csv("storm_data.csv")

# Select only the event type, fatalities/injuries, and damage
storm_data <- storm_data %>%
        select(EVTYPE, FATALITIES, INJURIES, 
               PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) %>%
        group_by(EVTYPE)
```

Next we want to find the total number of fatalities and injuries for each type 
of weather event. We will create two data sets with the top 25 most fatal and 
injurious weather events, and then re-combine them into one data set with an 
inner join.

```{r}
# Summarize data over max/total fatalities/injuries
storm_data2 <- storm_data %>%
        summarize(total_fatalities = sum(FATALITIES),
                  total_injuries = sum(INJURIES))

# Find most dangerous event types by fatalities and injuries
top = 25

top_fatal <- storm_data2 %>%
        arrange(desc(total_fatalities)) %>%
        head(top)

top_injury <- storm_data2 %>%
        arrange(desc(total_injuries)) %>%
        head(top)

# Combine data
most_harmful <- inner_join(top_fatal, 
                           top_injury, 
                           by = c("EVTYPE", 
                                  "total_fatalities", 
                                  "total_injuries"))
```

Finally we will filter our data to find the weather events that cause the most 
property damage, and the most damage to crops. We will only look for events that 
have caused damage in the billions of dollars (as denoted in the PROPDMGEXP and 
CROPDMGEXP columns), since those other events caused relatively little damage if
they only caused damage in the millions of dollars .

```{r}
# Filter data by property damage in the billions
storm_data3 <- storm_data %>%
        filter(PROPDMGEXP == "B") %>%
        summarize(prop_dmg = sum(PROPDMG)) %>%
        arrange(desc(prop_dmg))

# Filter by crop damage in the billions
storm_data4 <- storm_data %>%
        filter(CROPDMGEXP == "B") %>%
        summarize(crop_dmg = sum(CROPDMG)) %>%
        arrange(desc(crop_dmg))
```

## Results
Now we will plot the data. We will start by looking at the most harmful weather 
events.

```{r}
# Plot the most harmful weather events
most_harmful %>% gather(value = casualties, 
                        key = type, 
                        total_fatalities, 
                        total_injuries) %>% 
        ggplot(aes(x = reorder(EVTYPE, 
                               desc(casualties)),
                   y = log(casualties),
                   fill = type)) +
        geom_bar(stat = "identity",
                 position = "dodge") +
        labs(title = "Total Fatalities and Injuries by Weather Event",
             subtitle = paste("The Tornado event has caused",
                              "the most total injuries and total fatalities"),
             x = "Weather Event",
             y = "Log of the Total Number of Casualties") +
        theme(axis.text.x = element_text(angle = 45, 
                                         vjust = 1, 
                                         hjust=1),
              legend.title = element_blank(),
              legend.justification = c(1, 1),
              legend.position = c(1, 1),
              legend.background = element_rect(color = 1)) +
        scale_fill_discrete(breaks = c("total_fatalities",
                                       "total_injuries"),
                            labels = c("Total Fatalities",
                                       "Total Injuries"))
```

Next we will look at the weather events that have caused the most property 
damage.

```{r}
# Plot the most property damaging events
storm_data3 %>% head() %>%
        ggplot(aes(x = reorder(EVTYPE,
                               desc(prop_dmg)),
                   y = log(prop_dmg))) +
        geom_bar(stat = "identity") +
        theme(axis.text.x = element_text(angle = 45, 
                                         vjust = 1, 
                                         hjust=1)) +
        labs(title = "Billions of Dollars in Property Damage by Weather Event",
             subtitle = "Flooding has caused the most total property damage",
             x = "Weather Event",
             y = "Log of Total Amount of Property Damage (Billions of Dollars)")
```

Finally we will take a look at the events which have caused the most damage to 
crops.

```{r}
# Plot the most crop damaging events
storm_data4 %>% ggplot(aes(x = reorder(EVTYPE,
                                       desc(crop_dmg)),
                           y = crop_dmg)) +
        geom_bar(stat = "identity") +
        theme(axis.text.x = element_text(angle = 45, 
                                         vjust = 1, 
                                         hjust=1)) +
        labs(title = "Billions of Dollars in Crop Damage by Weather Event",
             subtitle = paste("Ice storms and river floods have", 
                              "casued the most total crop damage"),
             x = "Weather Event",
             y = "Total Amount of Crop Damage (Billions of Dollars)")
```
